% SPDX-License-Identifier: Apache-2.0
% Copyright (c) 2026 tomyimkc (Yim Kin Cheong), ORCID 0009-0005-9520-0033
%
% arXiv-ready source for the Sophia whitepaper (v0.11.0). Compiles with a standard
% TeX Live: pdflatex sophia-arxiv.tex (x2 for refs). Mirrors paper/sophia-whitepaper.md;
% the Markdown remains the canonical version. Honest framing: AGI-CANDIDATE, not proven AGI.
\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=black, urlcolor=blue, citecolor=blue}

\title{Sophia --- the Wisdom Gate:\\
A Provenance-Aware, Verifier-Gated Reasoning Layer\\
That Abstains Instead of Fabricating}
\author{Yim Kin Cheong\\
\small Independent researcher (tomyimkc), Hong Kong\\
\small ORCID: \href{https://orcid.org/0009-0005-9520-0033}{0009-0005-9520-0033}}
\date{Version 0.11.0 --- 2026-06-28}

\begin{document}
\maketitle

\begin{abstract}
Modern large language models (LLMs) confidently produce \textbf{attribution hallucinations}:
they credit the wrong author for an idea, merge distinct intellectual traditions, treat
legendary figures as literal authors, or cite authorities that do not exist --- and then reason
on top of those errors. We present \textbf{Sophia (the Wisdom Gate)}, a provenance-aware,
verifier-gated reasoning layer that \textbf{abstains instead of fabricating}. The control loop is
\emph{claim $\rightarrow$ verify against sources $\rightarrow$ accept $\cdot$ abstain $\cdot$
block}. Sophia couples (i) a queryable belief/provenance graph with confidence propagation,
retraction, and counterfactual analysis; (ii) deterministic and model-based verifiers that gate
claim promotion; (iii) calibrated abstention; and (iv) a fail-closed governance contract. We
evaluate under a \emph{pre-registered no-overclaim gate}: a number is \emph{validated} only with
$\ge 2$ independent judge families in consensus, reported inter-judge agreement (Cohen's
$\kappa$), $\ge 3$ runs, and confidence intervals. On a local 8B model, Sophia reduces
hallucinated attributions from $36.1\% \rightarrow 23.6\%$ ($\Delta\,12.5\%$, 95\% CI
$[5.6\%, 19.4\%]$) at $0\%$ false-positive cost; on ``I don't know'' traps a deterministic scorer
measures $0\%$ fabrication for the full pipeline versus $16.7$--$25\%$ for the raw model
($\kappa = 0.74$). This release adds a layered, \emph{independence-labelled} verification toolkit
(viral-claim fact-check, authorship and wrong-creator attribution checks, fabricated-citation
existence, and a multi-judge source-faithfulness check), and characterizes the limits of
open-world verification: a precision/recall ``no free lunch'' law, plus the discovery and
\emph{correction} of one of the system's own over-claims. We make \textbf{no} claim of general
intelligence.
\end{abstract}

\section{The problem}
LLMs are fluent before they are faithful. The failure mode we target is not generic
``hallucination'' but \textbf{attribution hallucination}, which is damaging because the error
becomes a \emph{premise} for downstream reasoning: crediting \emph{Confucius} with text from the
\emph{Dao De Jing}; attributing a 20th-century idea to an earlier figure (anachronism); treating
legendary figures as literal authors; in law, citing authorities that do not exist (the
\emph{Mata v.\ Avianca} failure) or misstating what a real authority held (the \emph{Ayinde}
misstated-authority failure). These are \emph{verifiable} errors --- there is ground truth a
machine can check --- which is precisely what a fluency-optimized decoder does not do by default.

\section{Design principles}
(1) Everything is a claim with sources. (2) Verification is first-class; the model is not trusted
to self-police. (3) Fail-closed, not fail-open: on low confidence, source conflict, or failed
verification, Sophia abstains or escalates --- it never fabricates to fill a gap.
(4) \emph{Functional} self-modeling, not consciousness --- no claim to subjective experience.
(5) Local-first and sovereign (multi-backend, airgap-capable).

\section{Method}
\textbf{The gate.} Retrieve $\rightarrow$ reason $\rightarrow$ verify $\rightarrow$
accept/abstain/block. \textbf{Belief \& provenance graph} with min-over-chain confidence,
retraction, and counterfactual removal (``what would I conclude if this source were withdrawn?'').
\textbf{Verifier classes}: deterministic (no model; e.g.\ citation existence, arithmetic) and
model-based (held to the no-overclaim gate). \textbf{Calibrated abstention} via self-consistency
selective prediction. \textbf{Fail-closed governance contract}: authorization $\rightarrow$
dataflow firewall $\rightarrow$ kill-switch $\rightarrow$ Bell--LaPadula classification
$\rightarrow$ taint-labeling on side-effecting tools, with re-verification of served output.

\subsection{The independent-verification toolkit (layered, independence-labelled)}
The verification step is a layer of independent oracles, each routed to the fabrication mode it
covers and each verdict tagged with its \textbf{independence tier}:
\begin{itemize}
  \item viral/general claims $\rightarrow$ professional fact-check verdicts (Google ClaimReview) --- \emph{high};
  \item authorship $\rightarrow$ Wikidata creator/author/discoverer records --- \emph{high};
  \item fabricated citations $\rightarrow$ study/DOI existence in Crossref --- \emph{high};
  \item wrong-creator attribution swaps $\rightarrow$ the Wikidata record vs the credited person --- \emph{high};
  \item misstated findings of a real source $\rightarrow$ a multi-judge entailment panel over an
        \emph{independent} retrieved source, strict-majority consensus --- \emph{medium} (model-based, flagged);
  \item everything else $\rightarrow$ \textbf{fail-closed abstention}.
\end{itemize}
Two commitments make this trustworthy rather than merely plausible: \textbf{fail-open on
ignorance} (a claim no oracle covers is not flagged --- never a fabricated contradiction), and
\textbf{honest coverage accounting} (a \emph{clean} verdict means ``no independent record
contradicts this,'' not ``true''). The result is a precision/recall trade governed by reference
quality, not a silver bullet.

\section{Measurement discipline: the no-overclaim gate}
Every public number must clear a pre-registered bar: $\ge 2$ independent judge families in
consensus (judge $\ne$ subject); reported inter-judge agreement ($\kappa \ge 0.40$); $\ge 3$ runs;
confidence intervals. Anything below is labeled \emph{illustrative} or \emph{candidate}.
Hidden-eval prompts are never published. \emph{Self-correction is part of the method}: an
independent judge audit revised one single-judge rate down; and an earlier $70.6\%$ over-block
figure was found to be a stale-report artifact and \textbf{withdrawn} (corrected to $5.9\%$).

\section{Results (selected)}
\begin{table}[h]\centering
\caption{Provenance delta (validated). Subject \texttt{dolphin-llama3:8b}; judges = consensus of
DeepSeek-Chat + Llama-3.3-70B (2 families); 3 runs.}
\begin{tabular}{lllll}
\toprule
Halluc.\ (alone) & Halluc.\ (gated) & $\Delta$ (95\% CI) & FP cost & Coverage \\
\midrule
$36.1\%$ & $23.6\%$ & $\mathbf{12.5\%\ [5.6\%, 19.4\%]}$ & $\mathbf{0.0\%}$ & $34.6\%$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Calibration / abstention (validated).} Full Sophia vs raw DeepSeek-Chat on
unknown-author/quote traps: $0\%$ fabrication for Sophia in all 3 runs vs $16.7$--$25\%$ raw;
calibration $\Delta\,22.0\%\ [14.5\%, 29.6\%]$; corroborated by GPT-4o + Claude-Sonnet ($\kappa = 0.74$).

\textbf{Verifier accuracy (deterministic).} \texttt{legal\_citation\_exists} catches fabricated
HK/UK/US citations (incl.\ the real \emph{Mata} fabrication) at $100\%$ on $N{=}14$ --- validating
the extraction + fail-closed logic, not a headline capability claim.

\textbf{Independent-verification toolkit (live, candidate).} The high-independence verifiers hold
$0\%$ clean over-block with no model judgment; the attribution-swap verifier (Wikidata), run with
separated answer/judge models over 3 runs with a bootstrap CI, caught $10.8\%\ [9.3\%, 11.6\%]$ of
a mixed pack ($\approx 2/3$ of the swap cases it targets) at $0.0\%\ [0,0]$ over-block. The honest
finding is a law, not a defect: catching open-world contamination requires \emph{either} an oracle
that covers the claim (sparse), \emph{or} fail-closed strictness (which over-blocks), \emph{or}
model knowledge (low independence). Trustworthiness comes from \emph{composing}
labelled-independence layers and abstaining.

\textbf{The honest tradeoff (candidate).} On a thin-source wiki corpus, strict grounding scores
$1.0$ vs $0.0$ on attribution-traps but collapses to $0.50$ vs $0.93$ on plain recall, so the raw
model wins overall; a typed-gate + graph-neighborhood hybrid recovers recall to $\approx 0.68$
while keeping all traps on the hard-abstain path.

\section{Related work}
Hallucination and faithfulness; selective prediction / calibration; retrieval-augmented
generation; fact-checking corpora (ClaimReview); structured knowledge bases (Wikidata, Crossref);
NLI/entailment; LLM-as-judge and its failure modes; defensive publication and provenance. (See the
project's prior-art survey for the full treatment.)

\section{Limitations}
Self-authored synthetic packs; one operator's keys; single-run live numbers are candidates pending
$\ge 3$-run CIs; oracle coverage is domain-bounded; model-based layers are lower-independence;
entity disambiguation bounds recall; \textbf{no third-party clean-room replication yet} --- the
reason \texttt{canClaimAGI} remains \textbf{false}.

\section{Reproducibility}
Keyless deterministic tests; live bench commands with documented env vars; a self-contained
third-party replication pack (\texttt{agi-proof/verification-replication/}: a two-tier runbook,
machine-readable expected results, a decontamination checklist, and a manifest checker). The
failure ledger and the trustworthiness capstone are first-class artifacts.

\section*{Availability and citation}
Source: \url{https://github.com/tomyimkc/sophia-agi} (Apache-2.0). Archived release with a
permanent DOI on Zenodo; cite via the repository's \texttt{CITATION.cff}.

\paragraph{Defensive-publication note.} This document is released to establish prior art and a
citable priority date for the method described. It is not a claim of AGI.

\end{document}
